---
title: "Top 50 Spotify Songs of 2019 - Exploratory Data Analysis"
author: "Rick Pohly"
date: "January 7, 2020"
output: github_document
pdf_document: default
html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**MOTIVATION**

As a musician, I'm always interested in what drives today's popular music. Since the music industry has experienced a massive paradigm shift towards streaming-based services over the last decade, a plethora of data are now available for all sorts of interesting analyses. I recognize that this dataset is tiny in scope (only 50 rows), but I chose it specifically for that reason, as a load-light way to practice, being a beginner in data science. I try to be careful not to draw any lofty conclusions given the virtually microscopic scale of this dataset. This is intended to be a very simple analysis for the purpose of getting the feel of EDA.

**Loading Data**

We begin by loading libraries and the wrangled top50.rda file into the object "topsongs" and converting the categorical genre variable to a factor.

```{r load and setup}
options(digits = 3)

# Load libraries and data
if(!require(ggplot2)){install.packages("ggplot2")}
if(!require(dplyr)){install.packages("dplyr")}
if(!require(RColorBrewer)){install.packages("RColorBrewer")}

load("rda/top50.rda")

# Convert categoricals to factors
topsongs <- topsongs %>% mutate_at(vars(genre), as.factor)
```

**Genre Grouping**

Having a closer look at the genre variable, we see that it contains 21 unique values. Upon inspection, it becomes clear that these are in fact subgenres. Since the dataset is only 50 rows, it might be wise to group these subgenres into the more general genres of pop, hip-hop, dance and latin for better stratification later.

```{r genre grouping}
length(levels(topsongs$genre))
levels(topsongs$genre)

# Group subgenres into more general genres
topsongs <- topsongs %>% 
  mutate(genre_group = as.factor(case_when(
    genre %in% c("atl hip hop", "escape room", "canadian hip hop", 
                 "dfw rap", "country rap") ~ "Hip-Hop",
    genre %in% c("brostep", "australian pop", "canadian pop",
                 "panamanian pop", "boy band", "electropop",
                 "pop") ~ "Pop",
    genre %in% c("latin", "r&b en espanol", "reggaeton", 
                 "reggaeton flow") ~ "Latin",
    genre %in% c("dance pop", "pop house", "trap music", 
                 "big room", "edm") ~ "Dance")))
```

**Summarization**

Looking at a summary of the numeric variables, we see that there is not a whole lot of variability for popularity. Values range from 70 to 95 with a mean value of 87.5 and a standard deviation of about 4.5. Most songs have a popularity value of 88 and 89. This is not unexpected. Since the dataset consists of the top 50 Spotify songs of 2019, it's safe to say that, by definition, *every* song is popular. Therefore, we will ignore this variable as a consequence of this lack of signal.

```{r summary}
# Summarize just the numeric data
summary(topsongs[,4:13])
sd(topsongs$popularity)

# Peek at popularity distribution
topsongs %>% ggplot(aes(x = popularity)) +
  theme_light() +
  geom_histogram(col = "black", binwidth = 1) +
  labs(x = "Popularity", y = "Song Count") +
  ggtitle("Popularity Distribution")
```

Stratifying by the generalized genre groups created earlier, we can gain some interesting insights.

```{r strat by genre}
# Which is the most common genre of the top 50?
topsongs %>% ggplot(aes(x = genre_group)) +
  theme_minimal() +
  geom_bar(aes(fill = genre_group)) +
  scale_fill_brewer(palette = "RdBu") +
  labs(x = "Genre", y = "Song Count") +
  ggtitle("Song Counts by Genre")

table(topsongs$genre_group)
```

First, and perhaps unsurprisingly, most of the songs in the dataset (17) fall into the category of pop, with dance a close a second, followed by latin and lastly, hip-hop. 

Next, let's have a look at deeper look. The energy, bpm and loudness variables are of some interest. Is there a connection between the amount of energy a song in a genre has and its speed or loudness? Is this a commonality among hit songs in general? 

```{r energy boxplot}
energybox.plt <- topsongs %>% 
  ggplot(aes(x = genre_group, y = energy)) +
  theme_light() +
  geom_boxplot(aes(fill = genre_group)) + 
  scale_fill_brewer(palette = "RdBu") +
  labs(x = "Genre", y = "Energy") +
  ggtitle("Energy by Genre")
energybox.plt

energydens.plt <- topsongs %>% 
  ggplot(aes(x = energy, fill = genre_group)) +
  theme_light() +
  geom_density(alpha = 0.4) + 
  scale_fill_brewer(palette = "RdBu") +
  labs(x = "Genre", y = "Energy") +
  ggtitle("Energy Density by Genre")
energydens.plt
```

It is clear from these plots that latin music has the highest median energy, and hip-hop the lowest. This could be a function of speed. We plot bpm against energy for a more fine-grained exploration.

```{r bpm vs energy}
bpmenergy.plt <- topsongs %>% group_by(genre_group) %>%
  mutate(rho = cor(energy, bpm)) %>%
  ggplot(aes(x = bpm, y = energy)) +
  theme_light() +
  geom_point() +
  facet_wrap(genre_group ~ rho) +
  labs(x = "Beats Per Minute", y = "Energy") +
  ggtitle("Stratified BPM vs Energy")
bpmenergy.plt
```

Keeping in mind that the data are very sparse, it appears that there is little to no correlation between beats per minute and energy level, with the noted exception of hip-hop. The rho value of hip-hop is around 0.57, showing a slightly strong positive correlation. Whether or not this is typical or simply due to lack of data remains to be seen.

Next, we have a look at loudness.

```{r energy vs loudness}
energyloud.plt <- topsongs %>%
  ggplot(aes(x = loudness, y = energy)) +
  theme_light() +
  geom_point(aes(color = genre_group), 
             alpha = 0.8) +
  scale_color_brewer(palette = "RdBu") +
  geom_smooth(method = "lm") +
  labs(x = "Loudness", y = "Energy") +
  ggtitle("Energy vs Loudness")
energyloud.plt
```

Even with such little data, we can instantly make out a suggestively strong positive correlation between energy and loudness, the strongest by far. A basic linear model seems to fit this quite well. The confidence interval is tightest at a loudness of about -6dB. Let's check the rho value to be sure.

```{r rho energy loudness}
cor(topsongs$loudness, topsongs$energy)
```

The value of rho for these two variables is 0.671, suggesting that loudness and energy are correlated. It is likely that if the dataset were larger, this rho value would be even higher. 

**CONCLUSION**

While most genres are of speeds and energy levels that span a larger range, it is latin music that has the highest energy along with the highest spoken word content. Hip-hop being a naturally spoken-word-based genre, this is a surprising find. One possible explanation is that hip-hop in recent years has shifted from the more rhythmic, spoken-word format of the 1980s and 1990s to the auto-tuned, melodic format that seems more prevalent today. In terms of predictive power, it is difficult to come to any trustworthy conclusion due to the small size of these data, but it is likely that loudness can be used to predict energy given a more comprehensive and sizeable dataset.